title: "SuppFig4B_Endothelial_Subclusters.Rmd"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
```

## Load integrated Seurat object

```{r load-object}
obj <- readRDS("lungQuery_After_AZ_Normalized.rds")
```

## Subset endothelial cells

```{r subset-endothelial}
endo <- subset(obj, subset = Level1 == "Endothelial")

endo
```

## Endothelial UMAP by cell subtype

```{r endothelial-umap-subtype, fig.width=8, fig.height=6}
DimPlot(
  endo,
  reduction = "umap",
  group.by = "new_celltype",
  label = TRUE,
  repel = TRUE
) +
  ggtitle("Endothelial Cell Lineage Subtypes")
```

## Endothelial UMAP by developmental group

```{r endothelial-umap-group, fig.width=8, fig.height=6}
DimPlot(
  endo,
  reduction = "umap",
  group.by = "Donor_Group"
) +
  ggtitle("Endothelial Cells Across Developmental Groups")
```

## Canonical endothelial lineage markers

```{r endothelial-featureplots, fig.width=12, fig.height=8}
FeaturePlot(
  endo,
  features = c("PECAM1", "VWF", "EMCN", "CLDN5", "ACKR1", "PLVAP", "CAR4", "RGCC"),
  reduction = "umap",
  ncol = 4
)
```

## Dot plot of endothelial markers by subtype

```{r endothelial-dotplot, fig.width=10, fig.height=6}
endo_markers <- c(
  "PECAM1", "VWF", "EMCN", "CLDN5",
  "ACKR1", "PLVAP", "CAR4", "RGCC"
)

DotPlot(
  endo,
  features = endo_markers,
  group.by = "new_celltype"
) +
  RotatedAxis() +
  ggtitle("Endothelial Lineage Marker Expression")
```

## Endothelial cell subtype proportion by group

```{r endothelial-proportions}
endo_prop <- endo@meta.data %>%
  count(Donor_Group, new_celltype) %>%
  group_by(Donor_Group) %>%
  mutate(Proportion = n / sum(n))

head(endo_prop)
```

```{r endothelial-proportion-plot, fig.width=10, fig.height=6}
ggplot(endo_prop, aes(x = Donor_Group, y = Proportion, fill = new_celltype)) +
  geom_bar(stat = "identity", position = "fill") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion of endothelial cells") +
  xlab("Developmental / disease group") +
  ggtitle("Endothelial Cell Lineage Composition Across Groups")
```

## Save endothelial metadata and proportions

```{r save-outputs}
write.csv(
  endo@meta.data,
  "Endothelial_Cell_Metadata.csv",
  row.names = TRUE
)

write.csv(
  endo_prop,
  "Endothelial_Cell_Proportions.csv",
  row.names = FALSE
)
```

## Session information

```{r session-info}
sessionInfo()
```
